### 2017-11-07 submission/manuscript.Rmd
bacground/intro edits

Colorectal cancer (CRC) is a growing world wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation in mouse models. There has also been numerous case control studies investigating the microbiota in both adenoma and carcinoma. Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. Many of the studies along with the current meta-analysis focus on identifying biomarkers or individual microbes but do not critically investigate the community role in the disease.

Using both fecal (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue samples (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] within our data analysis we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. We first assessed the diversity of controls, adenoma, and carcinoma individuals and tested whether they change and if it results in an increased relative risk of adenoma or carcinoma. Next, we assessed how this relative risk compared to CRC associated genera for both adenoma and carcinoma. Third, using Random Forest models we assessed whether the community context can increase the classification model area under the curve (AUC). Finally, we examine whether the studies that were used were adequately powered and if not what effect size they were powered for.

Our analysis found a continuous decrease in Shannon diversity from control to adenoma to carcinoma and a significantly increased relative risk for carcinoma with lower diversity. Using the CRC associated genera only this relative risk was higher than Shannon diversity. However, adding the community context in which these CRC associated genera are present increases prediction models AUC. Although we analyze a data set with a large number of total individuals each individual study was underpowered for effect size differences of 10% or below between the case and control.

>>>>

Colorectal cancer (CRC) is a growing world wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation in mouse models. There has also been numerous case/control studies investigating the microbiota in both adenoma and carcinoma. Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. Many of the studies along with the current meta-analysis focus on identifying biomarkers or individual microbes but do not critically investigate the community role in the disease.

Using both fecal (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue samples (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] we expand both the breadth and scope of the previous meta-analysis; to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes across disease (control to adenoma to carcinoma) and if it results in an increased relative risk for adenoma or carcinoma. Next, we assessed how typical CRC associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Third, using Random Forest models we analyzed whether the full community or only the CRC associated genera resulted in better model classification area under the curve (AUC). Finally, we examined at what effect size and sample n the studies that were used were adequately powered for.

Our analysis found a continuous decrease in Shannon diversity from control to adenoma to carcinoma and a significantly increased relative risk for carcinoma with this lower diversity. Using only the CRC associated genera this relative risk was higher than what was observed for Shannon diversity. However, by using the full community instead of only the CRC associated genera the AUC of the classification models increased. Although we analyze a data set with a large number of total individuals each individual study was underpowered for effect size differences of 10% or below between the case and control.

### 2017-11-07 submission/manuscript.tex
bacground/intro edits

Colorectal cancer (CRC) is a growing world wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in polyp formation in mouse models. There has also been
numerous case control studies investigating the microbiota in both
adenoma and carcinoma. Recently, a meta-analysis was published
investigating whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. Many of the studies along
with the current meta-analysis focus on identifying biomarkers or
individual microbes but do not critically investigate the community role
in the disease.

Using both fecal (total individuals = 1737) and tissue samples (total
samples = 492) totalling over 2229 total samples across 14 studies
{[}12--25{]} within our data analysis we expand both the breadth and
scope of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. We
first assessed the diversity of controls, adenoma, and carcinoma
individuals and tested whether they change and if it results in an
increased relative risk of adenoma or carcinoma. Next, we assessed how
this relative risk compared to CRC associated genera for both adenoma
and carcinoma. Third, using Random Forest models we assessed whether the
community context can increase the classification model area under the
curve (AUC). Finally, we examine whether the studies that were used were
adequately powered and if not what effect size they were powered for.

Our analysis found a continuous decrease in Shannon diversity from
control to adenoma to carcinoma and a significantly increased relative
risk for carcinoma with lower diversity. Using the CRC associated genera
only this relative risk was higher than Shannon diversity. However,
adding the community context in which these CRC associated genera are
present increases prediction models AUC. Although we analyze a data set
with a large number of total individuals each individual study was
underpowered for effect size differences of 10\% or below between the
case and control.

>>>>

Colorectal cancer (CRC) is a growing world wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in polyp formation in mouse models. There has also been
numerous case/control studies investigating the microbiota in both
adenoma and carcinoma. Recently, a meta-analysis was published
investigating whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. Many of the studies along
with the current meta-analysis focus on identifying biomarkers or
individual microbes but do not critically investigate the community role
in the disease.

Using both fecal (total individuals = 1737) and tissue samples (total
samples = 492) totalling over 2229 total samples across 14 studies
{[}12--25{]} we expand both the breadth and scope of the previous
meta-analysis; to investigate whether the bacterial community is an
important risk factor for both adenoma and carcinoma. To accomplish this
we first assessed whether the diversity changes across disease (control
to adenoma to carcinoma) and if it results in an increased relative risk
for adenoma or carcinoma. Next, we assessed how typical CRC associated
genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) affect the relative
risk of adenoma or carcinoma. Third, using Random Forest models we
analyzed whether the full community or only the CRC associated genera
resulted in better model classification area under the curve (AUC).
Finally, we examined at what effect size and sample n the studies that
were used were adequately powered for.

Our analysis found a continuous decrease in Shannon diversity from
control to adenoma to carcinoma and a significantly increased relative
risk for carcinoma with this lower diversity. Using only the CRC
associated genera this relative risk was higher than what was observed
for Shannon diversity. However, by using the full community instead of
only the CRC associated genera the AUC of the classification models
increased. Although we analyze a data set with a large number of total
individuals each individual study was underpowered for effect size
differences of 10\% or below between the case and control.
